\documentclass[11pt]{article}
\usepackage[margin=1.1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{microtype}
\hypersetup{colorlinks=true,linkcolor=blue!50!black,urlcolor=blue!50!black,citecolor=blue!50!black}

\title{GIST: A Gated Insight Synthesis Topology,\\
Native in a Coherence Calculus and Trainable as the\\
Recurrent State-Space/Convolutional Hybrid It Already Is}
\author{Edward Ellis\\
\small\texttt{github.com/ETEllis/bidi-coherence-delta-calculus} (substrate)}
\date{July 2, 2026}

\begin{document}
\maketitle

\begin{abstract}
GIST (Gated Insight Synthesis Topology) is a synthesis mechanism for
evidence-bearing reasoning: claims decompose into a 64-slot lattice
($2^6{=}4^3$), evidence superposes as phases, a corefold pyramid distills
each scope's triad to a singular conclusion, and a guarded
\emph{balanced-ternary commit} quantizes it---$+$ asserted, $-$ opposed,
$0$ a live open aperture---behind a barrier that holds contradiction-debt
rather than latching it. We present GIST built \emph{entirely} out of the
BiDi Coherence-Delta Calculus (CDC): the shipped mechanism is native
\texttt{.cdc} source executed by CDC's existing C micro-bridge with zero
host code of its own (machine-checked \texttt{host-debt 0}; 83/83 contract
expectations; 70 native witnesses). Pearl's do-operator requires no new
primitive: an intervention is a latched cell plus a severed inbound
channel, and we exhibit necessity, sufficiency-restoration, backdoor
severance, and frontdoor mediation as runtime-checked twin-field programs.
We then observe that the mechanism's forward dynamics already constitute a
neural architecture---a selective nonlinear state-space recurrence
(Kuramoto phase turn) over SO(2)-weighted graph convolutions with a ternary
bottleneck and hypercube pooling---and train that hybrid (109 parameters)
against labels produced by the native runtime itself. Initialized
\emph{at the calculus}, it reaches $0.984$ slot accuracy and $0.868$
global-verdict accuracy on held-out 64-slot scenes (majority baselines
$0.661/0.532$), and the trained weights export back to \texttt{.cdc},
where the same C runtime reproduces the model's conclusions on all 27
ternary inputs ($27/27$ parity). Learning itself is shown native: the
runtime now executes CDC's declared Hebbian plasticity as part of flow.
\end{abstract}

\section{Introduction: GIST before any neural network}

GIST is, at its conceptual base, a discipline for concluding under
evidence: (i)~\emph{scope} every question into a finite lattice of
micro-scopes; (ii)~\emph{superpose} evidence within each scope so that
agreement reinforces and contradiction cancels; (iii)~\emph{distill}
each scope's evidence triad through a fixed pyramid ($3{\to}2{\to}1$)
into a singular conclusion; (iv)~\emph{commit} conclusions only through
guards---a nonnegative-balance barrier over the committed word (leading
contradiction is debt and \emph{holds} with attribution rather than
latching) and an energy guard (a commit may not increase the scope's
variational potential); (v)~\emph{gate} any release of a synthesis on
coverage, parity, stability, counterfactual survival, causal lift,
global agreement, and retrieval saturation; and (vi)~\emph{reduce}
the lattice through a Hamming ladder whose reduction tree doubles as a
Merkle verification tree, ending in a council-reconciled, stamped
decision. The balanced-ternary middle value is load-bearing: $0$ is a
live crossing---``still open, contribute evidence''---not falsehood,
giving the mechanism a first-class state that language models
notoriously lack.

This paper reports GIST's construction \emph{from first principles}
inside an existing formal substrate, and the consequence that follows:
the mechanism is already a neural network, and can be trained as one
without leaving its own semantics.

\section{The substrate: BiDi Coherence-Delta Calculus}

CDC~\cite{bidi} is a native-language/kernel pair whose carrier is a
coherence algebra $\mathcal{C}=(C,\boxplus,\bot,\circlearrowright,
|{\cdot}|,\kappa)$ (standard model $C=\mathbb{R}^2$): superposition,
rotation, amplitude, and a commitment cosine $\kappa\in[-1,1]$ from which
the balanced trit is \emph{derived} with deadband $\delta=\tfrac12$:
$\tau=+1$ if $\kappa>\delta$, $-1$ if $\kappa<-\delta$, else $0$; boundary
openness $\varkappa=\max(0,1-|\kappa|/\delta)$ is maximal at the crossing.
Fields of phase-cells in bounded modules evolve by continuous \emph{flow}
(phase turns toward the afferent superposition) and discrete guarded
\emph{commit} (quantize, barrier, belief step, energy guard, latch), with
\emph{nest} exchanging child coherence upward and context downward. The
finite $n{=}6$ commit spectrum---$729$ words, $267$ admissible, $51$
localized (Motzkin), $20$ saturated, $5$ Catalan---is machine-checked and
grounds GIST's release tiers. The dyadic/triadic bridge
($2^6{=}4^3{=}64$) supplies GIST's address system.

GIST rides CDC \emph{subordinately}: its declaration files
(\texttt{gist.cdc}, laws, lattice, pyramid, counterfactuals, gate,
ladder, learning, neural, the training net, and the \texttt{.cdc}
datasets and learned weights) are executed and checked by CDC's own C
runtime and Python bootloader. The kernel contract asserts
\texttt{host-debt <= 0} and the verification gate confirms 83/83
expectations with 70 native witnesses. GIST is, to our knowledge, the
first CDC-first system: the calculus describing a mechanism which the
calculus's own bridge then executes, verifies, and (below) learns.

\section{The do-operator without a new primitive}

A CDC channel graph \emph{is} a structural causal model, so Pearl's
surgery~\cite{pearl} decomposes into two native acts: \emph{latch} the
intervened cell at its pole and \emph{sever} its inbound channels. The
counterfactual twin is a second field sharing initial conditions. Four
runtime-checked programs exhibit the ladder of causation natively:

\begin{itemize}
\item \textbf{Necessity} (\texttt{gist\_cf.cdc}): the factual field
commits the conclusion $+$; under $do(r_0{:=}-)$ the twin dissolves it to
the crossing---$r_0$ is load-bearing.
\item \textbf{Sufficiency-restoration}: a contested conclusion (word
\texttt{+-0}) is restored to $+$ by latching the opposing root.
\item \textbf{Backdoor severance}: a confounder co-moves $x$ and $y$
(word \texttt{+++}); $do(x{:=}-)$ with the backdoor severed yields only
\texttt{+-0}---the association overstates the causal effect.
\item \textbf{Frontdoor mediation}: with the confounder still supporting
$y$, $do(m{:=}-)$ on the mediator alone dissolves the conclusion,
isolating the mediated path.
\end{itemize}

Every expected phase and committed word in these files is checked by the
C runtime to $10^{-5}$. A companion reference implementation carries the
distributional layer (d-separation, do-calculus rules~1--3 as executable
rewrites, backdoor/frontdoor identification with honest refusal and
positivity attribution, exact abduction/PN/PS over the enumerable ternary
exogenous space) with $16/16$ suites passing, including exact agreement
between identified estimands and surgical ground truth across random
ternary SCMs.

\section{The hybrid: the network the mechanism already is}

Read as dynamics, GIST's forward pass is a neural architecture in the
precise, modern sense:

\begin{itemize}
\item \textbf{Recurrent nonlinear state space.} The flow step
$\Delta\theta_i=g\,\mathrm{Im}(A_ie^{-i\theta_i})$ (Kuramoto
turn~\cite{kuramoto}) is a recurrence whose coupling is state-dependent
---the selectivity that distinguishes modern SSMs~\cite{s4,mamba} arises
here from phase geometry itself.
\item \textbf{Convolutional filtration.} Two weight-shared graph
convolutions: the corefold pyramid inside every slot (six edge-type
filters), and a Hamming-1 lattice convolution on the $B$-cube (one SO(2)
filter per cube direction). Channels with weight $w$ and angle $\alpha$
are exactly the blocks $w\!\left[\begin{smallmatrix}\cos\alpha &
-\sin\alpha\\ \sin\alpha & \cos\alpha\end{smallmatrix}\right]$.
\item \textbf{Gating.} A learnable openness gate
$\sigma(\beta(|\kappa|-\delta))$ scales each cell's outbound influence:
decisive cells transmit, crossings attenuate---the calculus's aperture as
a trainable gate.
\item \textbf{Balanced-ternary bottleneck.} The commit quantizer with a
straight-through gradient; activations end as $\{-,0,+\}$ words behind
the same deadband the mechanism uses (cf.\ ternary-weight
LLMs~\cite{bitnet}, here as \emph{semantic} quantization).
\item \textbf{Ladder pooling.} $B$ hypercube contractions with a shared
alignment rotation per round pool $2^B$ slot singulars to one global
state.
\end{itemize}

\paragraph{Native supervision.} Training data is labeled by the
mechanism itself: scenes compile to \texttt{.cdc} batch programs
(evidence mass folded into channel weights, matching the runtime's
phase-only coupling), the C runtime commits them, and the committed words
are the labels---negatives via the mirror rule (the barrier is
assertion-leading, so ``no'' is a gated ``yes'' on the mirrored field).

\paragraph{Initialize at the calculus, train to anticipate.} The
provably-equivalent GIST-NN point (exact corefold weights, neutral
angles, unit gain) is the initialization; training then learns to
anticipate the \emph{converged} native reduction from $K{=}4$ recurrent
steps.

\begin{table}[t]
\centering
\caption{Held-out results, 64-slot lattice (109 parameters; numpy on one
CPU; 46 epochs chosen by early stopping). Majority baselines: slot
$0.661$, global $0.532$.}
\begin{tabular}{lccc}
\toprule
 & slot acc. & slot macro-F1 & global verdict acc.\\
\midrule
calculus init (untrained) & 0.785 & 0.644 & 0.540\\
trained & \textbf{0.984} & \textbf{0.980} & \textbf{0.868}\\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[t]
\centering
\caption{Ablations (8-slot lattice, same protocol).}
\begin{tabular}{lcc}
\toprule
variant & slot acc. & global acc.\\
\midrule
full hybrid, calculus init & 0.951 & \textbf{0.840}\\
random init & 0.968 & 0.750\\
no lattice convolution & \textbf{0.987} & 0.775\\
no openness gating & 0.984 & 0.745\\
no ternary bottleneck & 0.953 & 0.825\\
\bottomrule
\end{tabular}
\end{table}

The ablations localize the contributions: slot-level accuracy is carried
by the pyramid recurrence (removing lattice coupling even helps locally),
while the \emph{global verdict}---the judgment GIST exists to gate---is
best served by the full hybrid with calculus initialization; random
initialization and ablated gating cost $9$ and $9.5$ points of global
accuracy respectively.

\paragraph{Deployment closes the loop.} A parity-mode variant (gating
off, amplitude-blind, matching the C runtime's flow exactly) trains to
$0.985$ slot accuracy and exports its weights as a \texttt{.cdc} program;
\texttt{cdc\_native\_runtime infer} then reproduces the trained model's
singular conclusion on \emph{all} 27 ternary input words ($27/27$).
The same runtime that verifies the mechanism runs the trained model:
\texttt{./run.sh '+0-'}.

\paragraph{Training is native.} The runtime extension executes CDC's
declared plasticity: channels marked \texttt{plastic} follow phase
correlation through flow ($w\mathrel{+}=r(\cos(\theta_{src}{-}
\theta_{dst})-w)\,d$), witnessed by \texttt{gist\_learn.cdc} (a weak
channel learns an alignment, $0.10\to0.76$ in two steps, to $10^{-5}$).
On top of it the runtime's \texttt{train} command implements
clamped-teacher learning with no host code at all: the conclusion cell is
latched at the label pole (a do()-style clamp) while plasticity adapts the
weights through ordinary flow. Datasets are \texttt{.cdc} files whose
labels the runtime's own converged reduction computed; learned weights are
written back as \texttt{.cdc}. On the anticipation task (two flow layers
against a forty-step converged oracle), native training lifts held-out
accuracy from $0.813$ to $\mathbf{0.933}$ ($400$ training scenes, $10$
epochs, deterministic). The shipped repository trains, evaluates, and
infers entirely through the calculus and its C bridge; the gradient-based
hybrid results above are construction-tier reference experiments whose
harness deliberately stays out of the repository.

\section{Adapters and integration surfaces}

The reference tier exposes the mechanism to hosts: a one-call ternary
verdict bridge (\texttt{yes}/\texttt{no}/\texttt{maybe} $=$ the carrier,
with ``maybe'' returning the asks that would settle it, and ``no''
derived Pearl-symmetrically via the mirrored session), an MCP server, an
architecture-agnostic tool-call spec, and a reasoning-loop stitcher that
feeds a model's chain of thought through the lattice and injects the live
field state back into context. The reference engine is replay-verifiable
end-to-end (hash-chained ledgers, Merkle-rooted reductions, HKDF/HMAC
decision stamps) and carries a 100-test suite plus a 36-witness contract
of its own. A separate execution plan documents training the discipline
into open-weight LLMs with the native runtime as the (near-free) RL
reward function.

\section{Reproducibility and limitations}

Everything reported here reruns from the repository:
\texttt{native/verify\_gist.sh} (the whole native gate),
\texttt{torchcdc/train.py} (datasets, training, ablations, export,
parity; $\sim$6 minutes on one CPU), \texttt{torchcdc/test\_torchcdc.py},
and \texttt{./run.sh}. Determinism is by construction (hash-derived
streams; no wall clock in semantics). Limitations: the hybrid is small by
design (the point is the isomorphism and the closed loop, not scale);
the native runtime's flow is phase-only (amplitude semantics are folded
into weights at compilation, documented); identification tiers beyond
backdoor/frontdoor (Shpitser--Pearl ID) remain reference-tier work; and
the LLM translation is a designed protocol, not yet an executed run.

\begin{thebibliography}{9}
\bibitem{bidi} E.~Ellis. \emph{BiDi Coherence-Delta Calculus} (v0.2.4).
\url{https://github.com/ETEllis/bidi-coherence-delta-calculus}, 2026.
\bibitem{pearl} J.~Pearl. \emph{Causality: Models, Reasoning, and
Inference}. Cambridge University Press, 2nd ed., 2009.
\bibitem{kuramoto} Y.~Kuramoto. \emph{Chemical Oscillations, Waves, and
Turbulence}. Springer, 1984.
\bibitem{s4} A.~Gu, K.~Goel, C.~R\'e. Efficiently modeling long sequences
with structured state spaces. \emph{ICLR}, 2022.
\bibitem{mamba} A.~Gu, T.~Dao. Mamba: Linear-time sequence modeling with
selective state spaces. \emph{arXiv:2312.00752}, 2023.
\bibitem{bitnet} S.~Ma et al. The era of 1-bit LLMs: All large language
models are in 1.58 bits. \emph{arXiv:2402.17764}, 2024.
\end{thebibliography}

\end{document}
